---
title: "Explore and clean input data"
author: "Caitlin Mothes"
---

```{r}
library(tidyverse)
library(readxl)
library(sf)
library(mapview)
```

Read in datasets:

```{r}
sample <- read_xlsx("data/SAMPLE_geospatial_metadata_3.7.24.xlsx") %>% 
  janitor::clean_names()

genome <- read_xlsx("data/GENOME_geospatial_metadata.xlsx")

# Downloaded from here: https://www.eia.gov/maps/maps.php
play_basin <- read_sf("data/TightOil_ShaleGas_Plays_Lower48_EIA.shp") %>% 
  st_transform(crs = 4326) %>% 
  # remove spaces around hyphens for joining and change Denver basin name
  mutate(Shale_play = str_replace(Shale_play, " - ", "-"),
         Basin = str_replace(Basin, " Basin", ""))

sediment_basin <- read_sf("data/SedimentaryBasins_US_EIA.shp") %>% 
  mutate(NAME = str_to_title(NAME)) %>% 
  # need to transform all of these to WGS84
  st_transform(crs = 4326)

```


Read in international basins
```{r}
bowland <- read_sf("data/BGS_DECC_Bowland_study/topBHU_limit.shp") %>% 
  mutate(shale_play = "uk", shale_basin = "Bowland Shale") %>% 
  st_transform(4326)
mapview(bowland)


west_canada <- read_sf("data/wcsb_shapefiles/WCSBAtlasShp/fg0301_py_ll.shp")
mapview(west_canada)
```


Quick map of basins

```{r}
mapview(play_basin) +
  mapview(sediment_basin, color = "red")
```

Looks like sediment basins are just larger areas around play basins.

Check that all basins in sample datasets are in the basin shapefile

```{r}
sample %>% 
  filter(!basin %in% play_basin$Basin)

play_basin %>% 
  pull(Basin) %>% 
  unique()

sample %>% 
  pull(basin) %>% 
  unique()

genome %>% 
  pull(basin) %>% 
  unique()

# not an exact name match..also some basins in sample data not in shapefile
# no Bowland Shale, Sichuan, Western Canadian -> these are international
```

Some basins have separate polygons for formations, will need to join by matching sample formations

For the ones that do match, clean up basin names so we can join:

```{r}
# sample_clean <- sample %>%
#   # remove "Basin" from name and change Denver site name to match
#   mutate(Basin = str_remove_all(basin, "Basin") %>% 
#            str_replace("Denver-Julesburg", "Denver") %>% 
#            # trailing spaces were causing errors in join
#             str_remove_all(" "),
#          # capitalize formation names to match shapefile
#          Shale_play = str_to_title(formation) %>% 
#            # also change the formation wording for Anadarko sites
#            str_replace(fixed("Stack (Woodford)"), "Woodford"))

genome_clean <- genome %>% 
  # remove "Basin" from name and change Denver site name to match
  mutate(Basin = str_remove_all(basin, "Basin") %>% 
           str_replace("DJ", "Denver") %>% 
           # trailing spaces were causing errors in join
            str_remove_all(" ")) %>% 
  mutate(across(domain:species, ~str_remove(.x, ".*__")))

```

Now join both dataframes to shapefile (missing basin names will be dropped for now).

First create basin layer for app:

```{r}

play_basin_filtered <- play_basin %>%
  st_make_valid() %>% 
  group_by(Shale_play, Basin) %>% 
  summarise(st_union(geometry)) %>% 
  ungroup() %>% 
  filter(Shale_play %in% sample$shale_play) %>% 
  rename(shale_play = Shale_play, shale_basin = Basin)

#basin_sample <- 
  # trailing spaces causing errors in join
 # mutate(Shale_play = str_remove_all(Shale_play, " "))
  # mutate(
  #   Basin = str_remove_all(Basin, " ") %>%
  #     str_remove_all(., "Basin"),
  #   # fix typo
  #   Lithology = if_else(Lithology == "MIxed Shale & Chalk", "Mixed Shale & Chalk", Lithology)
  # ) %>%
  # # join to sample
  # inner_join(sample_clean, by = c("Basin", "Shale_play")) %>%
  # distinct(Basin, Shale_play, well_ID, Lithology, .keep_all = TRUE) %>%
  # group_by(Basin) %>%
  # mutate(n_wells = n()) %>%
  # select(Basin, Lithology, Shale_play, n_wells) %>%
  # # merge multipolygons to single one for basin/shale play
  # group_by(Basin, Shale_play, Lithology, n_wells) %>%
  # summarise(geometry = st_union(geometry)) %>%
  # ungroup() %>%
  # st_make_valid()

## not all basins are Shale

# make basin layer for genome map

basin_genome <- play_basin %>%
  # trailing spaces causing errors in join
  mutate(
    Basin = str_remove_all(Basin, " ") %>%
      str_remove_all(., "Basin"),
    # fix typo
    Lithology = if_else(Lithology == "MIxed Shale & Chalk", "Mixed Shale & Chalk", Lithology)
  ) %>%
   st_make_valid() %>% 
  group_by(Basin) %>% 
  summarise(geometry = st_union(geometry)) %>% 
  ungroup() %>% 
  st_make_valid() %>% 
  # join to sample
  inner_join(genome_clean, by = "Basin") %>%
  group_by(Basin) %>% 
  summarise(n_MAG_samples = n())
  




```

Create sample layer for app:

```{r}
sample_sf <- sample %>%
  left_join(play_basin_filtered, by = c("shale_play", "shale_basin")) %>%
  st_as_sf() %>%
  # convert to points
  st_centroid() %>%
  # edit lithology typo
  mutate(
    # make days_since_frack numeric
    days_since_frack = na_if(days_since_frack, "N/A|unk") %>%
      
      as.numeric(.),
    # edit time series stage for filtering
    timeseries_stage = if_else(timeseries_stage == "N/A" | timeseries == "no", "none", timeseries_stage))

# create variables for map symbology
sample_app <- sample_sf %>%
  st_drop_geometry() %>%
  group_by(well_id) %>%
  mutate(
   # n_samples = n(),
    max_days_since_frack = max(days_since_frack, na.rm = TRUE),
    min_days_since_frack = min(days_since_frack, na.rm = TRUE)
  ) %>%
  distinct(well_id, .keep_all = TRUE) %>%
  select(shale_play,
         well_id, 
         #n_samples,
         min_days_since_frack, 
         max_days_since_frack) %>%
  right_join(sample_sf, by = c("shale_play", "well_id")) %>%
  ungroup() %>%
  st_as_sf()

```

Create genome map data:

```{r}
# need to merge polygons per basin for genome data since no sub formation info
# play_basin_merge <- basin_app %>% 
#   st_make_valid() %>% 
#   group_by(Basin) %>% 
#   summarise(geometry = st_union(geometry)) %>% 
#   ungroup() %>% 
#   st_make_valid()

genome_app <- genome_clean #%>% 
  # inner_join(basin_genome, by = "Basin") %>% 
  # select(-Basin) %>% 
  # st_as_sf() %>% 
  # #convert to points
  # st_centroid()
```

Save output spatial dfs to read into app

```{r}
#save updated data
save(sample_app, play_basin, sediment_basin, file = "data/app_data_Update.RData")


#save(sample_app, genome_app, basin_sample, basin_genome, file = "data/app_data.RData")

# st_write(sample_summary, "data/sample_app.shp", append = FALSE)
# st_write(genome_sf, "data/genome_app.shp", append = FALSE)
```

Make time series plot with axis break

```{r}

fig_1 <- sample_app %>% 
 # filter(days)
    # mutate(Basin = factor(Basin, levels = c("WesternGulf", "Permian", "Appalachian", 
    #                                         "Anadarko", "Dener", "PowderRiver", "Michigan")),
    #        well_ID = fct_reorder(well_ID, max_days_since_frack)) %>%
 # arrange(Basin, well_ID, days_since_frack) %>% 
plotly::plot_ly(colors = "Dark2") %>% 
  add_trace(x = sample_app$days_since_frack,
            y = sample_app$well_id,
            split = ~sample_app$well_id,
            color = ~sample_app$shale_basin,
            name = ~sample_app$shale_basin,
            legendgroup = ~sample_app$shale_basin,
            type = 'scatter',
            mode = 'lines+markers',
            connectgaps = TRUE) %>% 
  layout(showlegend = TRUE, yaxis = list(type = "category",
  categoryorder = "array", categoryarray = sort(unique(sample_app$shale_basin)))) %>% 
         
  #categoryarray = sort(unique(sample_app$Basin))) %>% 
  # hacky way to get one trace per group in legend
  style(showlegend = FALSE, traces = c(1:4, 7:11, 13:16, 17, 21, 22:24, 26, 28))

fig_2 <-  sample_app %>% 
 # filter(days)
    # mutate(Basin = factor(Basin, levels = c("WesternGulf", "Permian", "Appalachian", 
    #                                         "Anadarko", "Dener", "PowderRiver", "Michigan")),
    #        well_ID = fct_reorder(well_ID, max_days_since_frack)) %>%
 # arrange(Basin, well_ID, days_since_frack) %>% 
plotly::plot_ly(colors = "Dark2") %>% 
  add_trace(x = sample_app$days_since_frack,
            y = sample_app$well_id,
            split = ~sample_app$well_id,
            color = ~sample_app$shale_basin,
            name = ~sample_app$shale_basin,
            legendgroup = ~sample_app$shale_basin,
            type = 'scatter',
            mode = 'lines+markers',
            connectgaps = TRUE) %>% 
  layout(showlegend = TRUE, yaxis = list(type = "category",
  categoryorder = "array", categoryarray = sort(unique(sample_app$shale_basin)))) %>% 
         
  #categoryarray = sort(unique(sample_app$Basin))) %>% 
  # hacky way to get one trace per group in legend
  style(showlegend = FALSE, traces = c(1:4, 7:11, 13:16, 17, 21, 22:24, 26, 28))

subplot(fig_1, fig_2, nrows = 1, shareY = TRUE, margin = 0.02) %>% 
  layout(xaxis = list(range = c(0, 2500)),
         xaxis2 = list(range = c(4500, 5000)))
  
```
