---
title: "Explore and clean input data"
author: "Caitlin Mothes"
---

```{r}
library(tidyverse)
library(readxl)
library(sf)
library(mapview)
```

Read in (test) datasets:

```{r}
sample <- read_xlsx("data/SAMPLE_geospatial_metadata.xlsx")

genome <- read_xlsx("data/GENOME_geospatial_metadata.xlsx")

# Downloaded from here: https://www.eia.gov/maps/maps.php
play_basin <- read_sf("data/ShalePlays_US_EIA_Dec2021.shp")

sediment_basin <- read_sf("data/SedimentaryBasins_US_May2011_v2.shp")
```

Quick map of basins

```{r}
mapview(play_basin) +
  mapview(sediment_basin, color = "red")
```

Looks like sediment basins are just larger areas around play basins.

Check that all basins in sample datasets are in the basin shapefile

```{r}
sample %>% 
  filter(!basin %in% play_basin$Basin)

play_basin %>% 
  pull(Basin) %>% 
  unique()

sample %>% 
  pull(basin) %>% 
  unique()

genome %>% 
  pull(basin) %>% 
  unique()

# not an exact name match..also some basins in sample data not in shapefile
# no Bowland Shale, Sichuan, Western Canadian -> these are international
```

Some basins have separate polygons for formations, will need to join by matching sample formations

For the ones that do match, clean up basin names so we can join:

```{r}
sample_clean <- sample %>%
  # remove "Basin" from name and change Denver site name to match
  mutate(Basin = str_remove_all(basin, "Basin") %>% 
           str_replace("Denver-Julesburg", "Denver") %>% 
           # trailing spaces were causing errors in join
            str_remove_all(" "),
         # capitalize formation names to match shapefile
         Shale_play = str_to_title(formation) %>% 
           # also change the formation wording for Anadarko sites
           str_replace(fixed("Stack (Woodford)"), "Woodford"))

genome_clean <- genome %>% 
  # remove "Basin" from name and change Denver site name to match
  mutate(Basin = str_remove_all(basin, "Basin") %>% 
           str_replace("DJ", "Denver") %>% 
           # trailing spaces were causing errors in join
            str_remove_all(" ")) %>% 
  mutate(across(domain:species, ~str_remove(.x, ".*__")))

```

Now join both dataframes to shapefile (missing basin names will be dropped for now).

First create basin layer for app:

```{r}

basin_sample <- play_basin %>%
  # trailing spaces causing errors in join
  mutate(
    Basin = str_remove_all(Basin, " ") %>%
      str_remove_all(., "Basin"),
    # fix typo
    Lithology = if_else(Lithology == "MIxed Shale & Chalk", "Mixed Shale & Chalk", Lithology)
  ) %>%
  # join to sample
  inner_join(sample_clean, by = c("Basin", "Shale_play")) %>%
  distinct(Basin, Shale_play, well_ID, Lithology, .keep_all = TRUE) %>%
  group_by(Basin) %>%
  mutate(n_wells = n()) %>%
  select(Basin, Lithology, Shale_play, n_wells) %>%
  # merge multipolygons to single one for basin/shale play
  group_by(Basin, Shale_play, Lithology, n_wells) %>%
  summarise(geometry = st_union(geometry)) %>%
  ungroup() %>%
  st_make_valid()

## not all basins are Shale

# make basin layer for genome map

basin_genome <- play_basin %>%
  # trailing spaces causing errors in join
  mutate(
    Basin = str_remove_all(Basin, " ") %>%
      str_remove_all(., "Basin"),
    # fix typo
    Lithology = if_else(Lithology == "MIxed Shale & Chalk", "Mixed Shale & Chalk", Lithology)
  ) %>%
   st_make_valid() %>% 
  group_by(Basin) %>% 
  summarise(geometry = st_union(geometry)) %>% 
  ungroup() %>% 
  st_make_valid() %>% 
  # join to sample
  inner_join(genome_clean, by = "Basin") %>%
  group_by(Basin) %>% 
  summarise(n_MAG_samples = n())
  




```

Create sample layer for app:

```{r}
sample_sf <- sample_clean %>%
  inner_join(basin_sample, by = c("Basin", "Shale_play")) %>%
  st_as_sf() %>%
  # convert to points
  st_centroid() %>%
  # edit lithology typo
  mutate(
    Lithology = if_else(Lithology == "MIxed Shale & Chalk", "Mixed Shale & Chalk", Lithology),
    days_since_frack = na_if(days_since_frack, "N/A") %>%
       # make days_since_frack numeric
       as.numeric(.)
  )

# create variables for map symbology
sample_app <- sample_sf %>%
  st_drop_geometry() %>%
  group_by(well_ID) %>%
  mutate(
    n_samples = n(),
    max_days_since_frack = max(days_since_frack, na.rm = TRUE)
  ) %>%
  distinct(well_ID, .keep_all = TRUE) %>%
  select(Basin, well_ID, n_samples, max_days_since_frack) %>%
  right_join(sample_sf, by = c("Basin", "well_ID")) %>%
  ungroup() %>%
  st_as_sf()

```

Create genome map data:

```{r}
# need to merge polygons per basin for genome data since no sub formation info
# play_basin_merge <- basin_app %>% 
#   st_make_valid() %>% 
#   group_by(Basin) %>% 
#   summarise(geometry = st_union(geometry)) %>% 
#   ungroup() %>% 
#   st_make_valid()

genome_app <- genome_clean #%>% 
  # inner_join(basin_genome, by = "Basin") %>% 
  # select(-Basin) %>% 
  # st_as_sf() %>% 
  # #convert to points
  # st_centroid()
```

Save output spatial dfs to read into app

```{r}
save(sample_app, genome_app, basin_sample, basin_genome, file = "data/app_data.RData")

# st_write(sample_summary, "data/sample_app.shp", append = FALSE)
# st_write(genome_sf, "data/genome_app.shp", append = FALSE)
```

Make time series plot

```{r}
sample_app %>% 
    mutate(Basin = factor(Basin, levels = c("WesternGulf", "Permian", "Appalachian", 
                                            "Anadarko", "Dener", "PowderRiver", "Michigan")),
           well_ID = fct_reorder(well_ID, max_days_since_frack)) %>%
 # arrange(Basin, well_ID, days_since_frack) %>% 

plotly::plot_ly(colors = "Dark2") %>% 
  add_trace(x = sample_app$days_since_frack,
            y = sample_app$well_ID,
            split = ~sample_app$well_ID,
            color = ~sample_app$Basin,
            name = ~sample_app$Basin,
            legendgroup = ~sample_app$Basin,
            type = 'scatter',
            mode = 'lines+markers',
            connectgaps = TRUE) %>% 
  layout(showlegend = TRUE, yaxis = list(type = "category",
  categoryorder = "array", categoryarray = sort(unique(sample_app$Basin)))) %>% 
         
  #categoryarray = sort(unique(sample_app$Basin))) %>% 
  # hacky way to get one trace per group in legend
  style(showlegend = FALSE, traces = c(1:4, 6:10, 12:14, 16, 18, 20:21, 23:24))
  
```
