---
title: "Explore and clean input data"
author: "Caitlin Mothes"
---

```{r}
library(tidyverse)
library(readxl)
library(sf)
library(mapview)
```

Read in (test) datasets:

```{r}
sample <- read_xlsx("data/SAMPLE_geospatial_metadata.xlsx")

genome <- read_xlsx("data/GENOME_geospatial_metadata.xlsx")

# Downloaded from here: https://www.eia.gov/maps/maps.php
play_basin <- read_sf("data/ShalePlays_US_EIA_Dec2021.shp")

sediment_basin <- read_sf("data/SedimentaryBasins_US_May2011_v2.shp")
```

Quick map of basins

```{r}
mapview(play_basin) +
  mapview(sediment_basin, color = "red")
```

Looks like sediment basins are just larger areas around play basins.

Check that all basins in sample datasets are in the basin shapefile

```{r}
sample %>% 
  filter(!basin %in% play_basin$Basin)

play_basin %>% 
  pull(Basin) %>% 
  unique()

sample %>% 
  pull(basin) %>% 
  unique()

genome %>% 
  pull(basin) %>% 
  unique()

# not an exact name match..also some basins in sample data not in shapefile
# no Bowland Shale, Sichuan, Western Canadian, DJ? Does this stand for Denver-Julesberg
```

Some basins have separate polygons (e.g., Permian), can subset by formation/sediment play

For the ones that do match, clean up basin names so we can join

```{r}
sample_clean <- sample %>%
  # remove "Basin" from name and change Denver site name to match
  mutate(Basin = str_remove_all(basin, "Basin") %>% 
           str_replace("Denver-Julesburg", "Denver") %>% 
           # trailing spaces were causing errors in join
            str_remove_all(" "),
         # capitalize formation names to match shapefile
         Shale_play = str_to_title(formation) %>% 
           # also change the formation wording for Anadarko sites
           str_replace(fixed("Stack (Woodford)"), "Woodford"))

genome_clean <- genome %>% 
  # remove "Basin" from name and change Denver site name to match
  mutate(Basin = str_remove_all(basin, "Basin") %>% 
           str_replace("DJ", "Denver") %>% 
           # trailing spaces were causing errors in join
            str_remove_all(" ")) %>% 
  mutate(across(domain:species, ~str_remove(.x, ".*__")))

```

Now join both dataframes to shapefile (missing basin names will be dropped for now)

```{r}
# trailing spaces causing errors in join
play_basin <- play_basin %>%
  mutate(Basin = str_remove_all(Basin, " ") %>%
           str_remove_all(., "Basin"))
  

sample_sf <- sample_clean %>% 
  inner_join(play_basin, by = c("Basin", "Shale_play")) %>% 
  st_as_sf() %>% 
  # convert to points
  st_centroid() %>% 
  # edit lithology typo
  mutate(Lithology = if_else(Lithology == "MIxed Shale & Chalk", "Mixed Shale & Chalk", Lithology))


```

```{r}
# need to merge polygons per basin for genome data since no sub formation info
play_basin_merge <- play_basin %>% 
  st_make_valid() %>% 
  group_by(Basin) %>% 
  summarise(geometry = st_union(geometry)) %>% 
  ungroup()

genome_app <- genome_clean %>% 
  inner_join(play_basin_merge, by = "Basin") %>% 
  st_as_sf() %>% 
  #convert to points
  st_centroid()
```


Create variables for map symbology

```{r}
# calculate # wells/basin and #samples/well
sample_app <- sample_sf %>%
  st_drop_geometry() %>%
  # replace all N/A
  mutate(across(where(is.character), ~ na_if(., "N/A")),
         # make days_since_frack numeric
         days_since_frack = as.numeric(days_since_frack)) %>%
  group_by(well_ID) %>%
  mutate(
    n_samples = n(),
    max_days_since_frack = max(days_since_frack, na.rm = TRUE)
  ) %>%
  distinct(well_ID, .keep_all = TRUE) %>%
  group_by(Basin) %>%
  mutate(n_wells = n()) %>%
  select(Basin, well_ID, n_wells, n_samples, max_days_since_frack) %>%
  right_join(sample_sf, by = c("Basin", "well_ID")) %>% 
  ungroup() %>% 
  st_as_sf()


```


Save output spatial dfs to read into app

```{r}
save(sample_app, genome_app, file = "data/app_data.RData")

# st_write(sample_summary, "data/sample_app.shp", append = FALSE)
# st_write(genome_sf, "data/genome_app.shp", append = FALSE)
```


